Hawaiian reef communities are more distinct across a 90 m depth gradient than 2,500 km of shallow habitat

### 2024-03-18 manuscript.Rmd
two rounds of comments from RJT

---
title: "Depth over distance: Hawaiian cryptic reef communities are more distinct across ~90 m depth than ~2500 km of shallow habitat"
author:
  - Kaleonani K.C. Hurley:
      institute:
        - himb
        - nerr
      equal_contributor: yes
      correspondence: yes
      email: "hurleyk@hawaii.edu"
  - Mykle L. Hoban:
      institute:
        - himb
      equal_contributor: yes
  - Molly A. Timmers:
      institute:
        - himb
        - natgeo
  - Kerry Reardon:
      institute: noaa
  - Robert J. Toonen:
      institute: himb
institute:
  - himb: Hawai&#x02BB;i Institute of Marine Biology
  - nerr: He&#x02BB;eia National Estuarine Research Reserve
  - natgeo: National Geographic Pristine Seas
  - noaa: National Oceanic and Atmospheric Administration
abstract:
  Shallow coral reefs are extensively studied, and although scleractinian corals have been recorded to 165 m, little is known about other mesophotic coral reef ecosystem (MCE) inhabitants. Brachyuran crabs fill many ecological and trophic niches on reefs, making them ideal candidates for evaluating species composition among depths. Here we ask if MCEs host the same communities as the shallower reefs spread across the entire 2400 km Hawaiian Archipelago. We deployed Autonomous Reef Monitoring Structures (ARMS) for two years among shallow sites (12 m) across the latitudinal gradient of the Hawaiian Archipelago to compare directly with a depth gradient (12, 30, 60, and 90 m) south of Oʻahu island. A total of 2284 individual crabs representing 90 morphospecies (19 families) were found. Assemblage composition was not statistically significantly different among shallow sites spread across the archipelago but was significantly different and highly stratified across the depth gradient.  We show that 90 m of depth is a more potent determinant of brachyuran community composition than the latitudinal, geographic (high vs low islands), or anthropogenic impact (human population size) gradient across the archipelago. Deeper reefs host significantly different brachyuran communities than shallow ones in Hawai&#x02BB;i, and only 3 of 90 morphospecies (~3%) occurring across the entire depth range (mesophotic to shallow) were sampled. Further research on the cryptic reef-dwelling fauna is needed for identifying areas of unique biodiversity, as well as boundaries between depth-stratified reef ecosystems.
bibliography: references/citations.bib
csl: references/style.csl
output:
  word_document:
    pandoc_args:
      - "--toc-depth=2"
      - "--filter=resources/pandoc-crossref"
      - "--lua-filter=resources/scholarly-metadata.lua"
      - "--lua-filter=resources/author-info-blocks.lua"
      - "--reference-doc=resources/ref.docx"
    keep_md: true
tblPrefix: table
figPrefix:
  - "figure"
  - "figures"
chapters: True
chaptersDepth: 1
chapDelim: ""
header-includes:
  - \usepackage{bm}
  - \usepackage{upgreek}
---
<!--word count: Proceedings B publishes articles up to 12 pages, including figures and tables, with no more than 10 000 words. The word count includes the title page, author names, abstract, figure legends, tables and the reference list. As a rough guide, 10 000 words = 12 pages without additional figures; 850 words = 1 page and 2-4 small figures = 1 page. For example, an acceptable paper could be made up of 9 000 words and 4 small figures or 8500 words, 2 large figures and 4 small figures.-->
### Keywords {-}
<div custom-style="mono">ARMs, mesophotic coral ecosystems, community structure, dispersal, Hawai&#x02BB;i, brachyuran crabs</div>

@@

print_table <- function(tbl,id="",caption="",hdr_case="sentence",hide=character(0),accuracy = 0.01) {
  if (!grepl("^: ",caption)) {
    caption <- paste0(": ",caption)
  }

@@

  tbl <- tbl %>%
    select(-any_of(hide)) %>%
    mutate(
      across(
        where(is.numeric),
        ~num(.x,accuracy)
      )
    )

  # can also use rename_with(str_to_sentence)
  names(tbl) <- switch(
    str_to_lower(hdr_case),
    sentence = str_to_sentence(names(tbl)),
    title = str_to_title(names(tbl)),
    names(tbl)
  )

@@

# palette for island/group colors
# island_pal <- c(
#   "Hawaii Island" = "#8a3ffc",
#   "Hawai‘i Island" = "#8a3ffc",
#   "Maui" = "#007d79",
#   "Oahu" = "#ff7eb6",
#   "O‘ahu" = "#ff7eb6",
#   "Kauai" = "#fa4d56",
#   "Kaua‘i" = "#fa4d56",
# 
#   "French Frigate Shoals" = "#fff1f1",
#   "Lalo" = "#fff1f1",
#   "Lalo (French Frigate Shoals)" = "#fff1f1",
#   "Lisianski" = "#6fdc8c",
#   "Kapou" = "#6fdc8c",
#   "Kapou (Lisianski)" = "#6fdc8c",
#   "Pearl and Hermes Reef" = "#4589ff",
#   "Manawai" = "#4589ff",
#   "Manawai (Pearl and Hermes Reef)" = "#4589ff",
#   "Kure Atoll" = "#d12771",
#   "Hōlanikū" = "#d12771",
#   "Hōlanikū (Kure Atoll)" = "#d12771"
# )
island_pal <- c(
  "Hawaii Island" = "#5E4FA2",
  "Hawai‘i Island" = "#5E4FA2",
  "Maui" = "#3288BD",
  "Oahu" = "#66C2A5",
  "O‘ahu" = "#66C2A5",
  "Kauai" = "#ABDDA4",
  "Kaua‘i" = "#ABDDA4",

@@

The structure and composition of coral reef communities is influenced by a suite of environmental and anthropogenic factors, yet the nature and makeup of those factors remain poorly resolved. Coral reefs are among the most diverse habitats in the world, and they face increasing threats of degradation, including climate-driven bleaching and phase shifts to alternate ecological regimes [@grahamManagingResilienceReverse2013; @norstromAlternativeStatesCoral2009; @pandolfiGlobalTrajectoriesLongTerm2003]. In the face of these and other threats, understanding drivers of community structure is critically important [@hughesRisingChallengeSustaining2010; @pageSeekingResistanceCoral2019]. To date, the majority of efforts in this area have focused on conspicuous taxa such as corals and fishes [e.g., @benfieldComparisonCoralReef2008; @caballero-aragonWaveExposureTemperature2023; @kaneTrophicDesignationLive2017; @langeWaveExposureShapes2021; @vanlierHabitatConnectivityComplexity2018] or microbes [e.g., @desouzaImportanceDepthTemperature2023; @morrowEnvironmentalDriversMicrobial2016]. However, these taxa make up a minority of coral reef biodiversity and thus may not be indicative of community patterns overall [[cite?]]{.mark}. Here, we investigate the factors influencing the community structure of a group of cryptic coral reef taxa across the Hawaiian Archipelago.

Much of our understanding of biotic zonation and community structure in the ocean originates with classic studies from the intertidal [e.g., @connellInfluenceInterspecificCompetition1961; @connellEffectsCompetitionPredation1961; @paineIntertidalCommunityStructure1974]. Such studies point to a combination of physical and biological factors<!--, including phyiological tolerance [@dennyPhysicalProcessesThat2001; @someroThermalPhysiologyVertical2002] and species interactions [@brunoHabitatModificationFacilitation2001; @mengeRockyIntertidalCommunities2001], that contribute to the structure of intertidal communities. Although there are exceptions [e.g., @woottonIndirectEffectsPrey1992; @bertnessTestingRelativeContribution1999]-->, but (with exceptions) a generalization has emerged <!--for the intertidal -->that communities' lower bathymetric limits are regulated by ecological interactions whereas upper limits are determined by abiotic factors. However, these models describe zonation at individual localities rather than community variation over larger spatial scales and may not apply to coral reefs, which are subject to more complex community dynamics [@abramsRecruitmentLotteriesCoexistence1984; @bussCompetitiveNetworksNontransitive1979; @hixonPredationPreyRefuges1993; @hubbellUnifiedNeutralTheory2011; @moraEcologyFishesCoral2015; @saleStockRecruitRelationshipsRegional1982; @sheppardBiologyCoralReefs2017]. In any case, it is likely that while ecological interactions are important for structuring communities at the local scale, external factors may exert stronger influence across greater distances [[cite?]]{.mark}.

The Hawaiian Archipelago comprises high volcanic islands and atolls stretching across ~2,500 km in the North Pacific Ocean, and is typically regarded as two regions: the main Hawaiian Islands (MHI), which include the eight inhabited high islands, and the Northwestern Hawaiian Islands (NWHI), the uninhabited atolls and banks northwest of Kaua&#x02BB;i and Ni&#x02BB;ihau that make up the Papahānaumokuākea Marine National Monument. The islands are subject to anthropogenic and physical gradients across their span [@selkoeMapHumanImpacts2009], which we hypothesize to influence coral reef communities. Studies examining community structure across environmental gradients have been criticized for confounding multiple potentially covarying factors (such as habitat area and/or complexity) across the measured gradients [e.g., @bellwoodRegionalScaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007a]. Thus, we used autonomous reef monitoring structures (ARMS)---standardized collection devices designed to mimic reef structural complexity---to sample cryptic communities while controlling for habitat<!--differences in habitat variables--> [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009]. Specifically, we used ARMS to sample brachyuran crabs, a group comprising a notable proportion of the coral reef cryptofauna [289 species in Hawai&#x02BB;i, @castroCatalogAnomuranBrachyuran2011]. We compared crab communities among islands and regions (MHI vs. NWHI) and examined the effects of a suite of possible environmental and anthropogenic drivers.

A likely important but under-investigated aspect of coral reef community structure is depth. Most studies to date have focused on the upper ~30 m of the water column, yet coral reefs and associated communities are known to extend beyond 150 m [@pyleExploringDeepCoral1996; @pyleMesophoticCoralEcosystems2019], with photosymbiotic corals reported to 170 m [@rouzeSymbioticAssociationsDeepest2021]. As much as two thirds of coral reef ecosystems lie in this under-explored zone [@hindersteinThemeSectionMesophotic2010]. These habitats, known as mesophotic coral ecosystems (MCEs), are typically defined as those extending from ~30--150 m depth [@hindersteinThemeSectionMesophotic2010; @kahngCommunityEcologyMesophotic2010]. <!--and were believed to experience dampened disturbances in comparison to shallow reefs and could therefore act as deep water refugia for shallow reef species threatened by anthropogenic impacts --> They have been hypothesized as possible refugia from disturbances on shallow reefs [@glynnCoralReefBleaching1996; @bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although studies have found evidence that deep reefs are unique and uniquely threatened, challenging this notion [@diazLightTemperatureDrive; @diazMesophoticCoralBleaching2023; @hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018]. <!--However, recent literature describing species compositions along depth gradients reveal a more complex picture among taxa and locations because some groups may use mesophotic reefs as a refuge in some places, while others show community turnover with depth such that MCE habitats are ecologically distinct from adjacent shallow water coral reefs [@hurleyAssessmentShallowMesophotic2016; @smithCaribbeanMesophoticCoral2016; @semmlerAreMesophoticCoral2017; @bongaertsDeepReefsAre2017; @rochaMesophoticCoralEcosystems2018]. Further, there are a range of connections depending on the taxa – large, mobile organisms like fish tend to have more shared species across depth than sedentary organisms such as corals and algae, but even among the fishes the community structure of MCEs is often distinct [@kaneHighLevelsMesophotic2014; @pyleComprehensiveInvestigationMesophotic2016 ; @asherMesophoticDepthGradients2017; @fukunagaChangesMesophoticReef2017; @colemanShiftingReefFish2018; @pinheiroDeepReefFishes2019]. Although the majority of coral reef biodiversity consists of the algal and invertebrate cryptobenthic organisms living on and within the reef structure [@fautinOverviewMarineBiodiversity2010; @knowltonCoralReefBiodiversity2010; @fisherSpeciesRichnessCoral2015], the studies listed above focus almost exclusively on the corals and fishes which are more easily surveyed.-->Given the strong reported differences between deep and shallow reefs across taxa and the paucity of research on drivers of cryptic coral reef community variation, we sought to answer the following research questions: (i) How do shallow crab communities vary along the natural and anthropogenic gradients of the Hawaiian Archipelago? (ii) Do shallow crab communities differ between the Main and Northwestern Hawaiian Islands? (iii) What environmental and/or anthropogenic factors drive crab community structure in shallow coral reefs across the Hawaiian Archipelago? (iv) What are the relative contributions of depth (shallow vs. mesophotic) and island region to crab community variation across the Hawaiian Archipelago?

<!-- Here, we employed autonomous reef monitoring structures (ARMS)  Using ARMS, we sampled brachyuran crab communities across the Hawaiian Archipelago to test the effects of environmental and anthropogenic drivers on the structure of cryptic reef communities.    Given the number and magnitude of differences between these regions, we hypothesized that species diversity would be different between regions (MHI vs NWHI). We therefore looked at community composition along the latitudinal gradient and investigated community composition in response to physical gradients (depth, temperature), biological gradients (coral cover, chlorophyll concentration), geological gradients (island slope, distances between neighboring islands), and anthropogenic gradients (human population count, human impacts). We hypothesized that the communities found across the latitudinal gradient encompassing all of these environmental variables would show fewer differences between individual islands or atolls, but would be greatest between the two regions, MHI and NWHI.  Finally, to investigate whether a latitudinal gradient or depth gradient would be a stronger force in driving community composition, we compared the reef crab assemblage from across the shallow reefs in the Hawaiian Archipelago to reef crab assemblages across a mesophotic depth gradient.  -->

<!-- The Hawaiian Archipelago is a volcanic island arc made up of high islands and atolls spanning ~9 degrees of latitude and ~2,500 km. It is typically regarded as two regions: the inhabited main Hawaiian Islands (MHI), which include the eight islands from Hawai&#x02BB;i Island to Kauai&#x02BB;i, and the largely uninhabited Northwestern Hawaiian Islands (NWHI), which constitute the 10 low islands and atolls to the northwest of Kaua&#x02BB;i. We compared shallow reef crab communities across the 2500 km archipelago and  we compared reef crab assemblages from the shallow reefs to the reef crab assemblages across a mesophotic depth gradient off of Kewalo, Oʻahu. By controlling for habitat area and complexity and by incorporating sites across various environmental and human impact gradients, we sought to better understand whether latitude, the environment, or depth is more influential in structuring reef crab communities. -->

@@

ARMS units were retrieved by encapsulating them in-situ in a 100 μm mesh-lined crate to prevent escape of motile organisms. Once returned to the surface, they were placed in containers of aerated, filtered seawater, crates were removed, and units were systematically disassembled plate-by-plate. Seawater surrounding disassembled units was sieved through 2 mm, 500 μm, and 100 μm mesh geologic sieves [see @lerayDNABarcodingMetabarcoding2015 for processing details], and retained macroorganisms were preserved in 95% ethanol. All adult brachyuran crabs ($\geq$ 5 mm) were photographed, identified (to species where possible, otherwise morphospecies), and preserved in 95% ethanol. Specimens were collected under permits [#xxx]{.mark} (NWHI) and [#xxx]{.mark} (MHI) and accessioned into the invertebrate zoology collections at the Florida Museum of Natural History ([catalog numbers xxx--xxx]{.mark}).

@@

To examine patterns of crab diversity across the islands, we used `estimate_richness` to calculate species richness and Simpson diversity (Simpson = $1-\sum_{i=1}^{R}p_i^2$, where R is the total number of species, and $\mathrm{p}_i$ is the proportional abundance of each species). We used nonparametric Kruskal-Wallis ranked sum tests to compare diversity between shallow and mesophotic samples and between island groups (MHI vs. NWHI). Island group comparisons were performed both with and without mesophotic samples. <!--To explore environmental influences on alpha diversity, we used a linear mixed-effect model with Simpson or richness as response variables, the environmental variables listed below as fixed effects, and island nested within island group (MHI vs NWHI) as random effects. Models were generated using the `lmer` function from the `lme4` package [@batesFittingLinearMixedEffects2015]. -->

@@

We assessed variation among groups using `betadisper` and `adonis2`. These functions implement analysis of multivariate homogeneity of group dispersions (PERMDISP) and multivariate permutational analysis of variance (PERMANOVA) respectively. PERMDISP is used to assess variability within groups and PERMANOVA is used to assess differences among groups. Significance was assessed with permutation tests in both cases. Variation among sampling groups (island group and depth) was further examined and visualized using canonical analysis of principal coordinates (CAP) based on discriminant analysis [@andersonCanonicalAnalysisPrincipal2003] as implemented in the `CAPdiscrim` function in `BiodiversityR` [@kindtTreeDiversityAnalysis2005].

@@

The `r nsamples(cc$crabs_untransformed)` ARMS units deployed across the Hawaiian Islands (in `r format_range(cc$crab_data$depth)` m depth) yielded a total of `r num(sum(sample_sums(cc$crabs_untransformed)),1)` adult crabs. Units contained `r num(all_summary$abundance_mean,0.1)` $\pm$ `r num(all_summary$abundance_sd,0.1)` (mean $\pm$ SD) individuals, and abundance ranged from `r format_range(all_summary$abundance_range)`. Crabs were identified to `r sum(unlist(all_summary$types_breakdown))` distinct types (`r all_summary$types_breakdown$species` species and `r all_summary$types_breakdown$morphospecies` morphospecies) in `r all_summary$taxonomy_breakdown$genera` genera and `r all_summary$taxonomy_breakdown$families` families, which accounted for ~36% of the known crab diversity reported from Hawaiian waters [@castroCatalogAnomuranBrachyuran2011]. The four most abundant species were: `r with(all_summary,top_list(top,top_n,top_units,fig="fig_crabs"))`. `r str_to_sentence(english(all_summary$singleton_count))` species occurred in a single unit across all sites (singletons), accounting for ~`r num(all_summary$singleton_count/ntaxa(cc$crabs_untransformed)*100,1)`% of total recovered crab species.

@@

The `r nsamples(cc$crabs_deep)` units deployed at mesophotic depths $\geq$ 30 m on O&#x02BB;ahu yielded fewer species and individuals compared to the complete dataset, with `r num(sum(sample_sums(cc$crabs_deep)),1)` crabs overall, `r sum(unlist(deep_summary$types_breakdown))` distinct types (`r deep_summary$types_breakdown$species` species and `r deep_summary$types_breakdown$morphospecies` morphospecies), `r num(deep_summary$abundance_mean,0.1)` $\pm$ `r num(deep_summary$abundance_sd,0.1)` individuals per unit, and abundance ranging from `r format_range(deep_summary$abundance_range)`. The four most abundant species in the deep ARMS did not overlap with those from the complete dataset: `r with(deep_summary,top_list(top,top_n,top_units))`. There were `r deep_summary$singleton_count` singletons, which proportionally accounted for ~`r num(deep_summary$singleton_count/ntaxa(cc$crabs_deep)*100,1)`% of species found on deep ARMS.

@@

<!-- ### Effects of environmental variables {-} -->

## Brachyuran community structure {-}

>>>>

---
title: "Hawaiian reef communities are more distinct across a 90 m depth gradient than 2,500 km of shallow habitat"
author:
  - Kaleonani K.C. Hurley:
      institute:
        - himb
        - nerr
      equal_contributor: yes
      correspondence: yes
      email: "hurleyk@hawaii.edu"
  - Mykle L. Hoban:
      institute:
        - himb
      equal_contributor: yes
  - Molly A. Timmers:
      institute:
        - himb
        - natgeo
  - Derek J. Skillings:
      institute: 
        - uncg
  - Kerry Reardon:
      institute: noaa
  - Robert J. Toonen:
      institute: himb
institute:
  - himb: Hawai&#x02BB;i Institute of Marine Biology
  - nerr: He&#x02BB;eia National Estuarine Research Reserve
  - natgeo: National Geographic Pristine Seas
  - uncg: 'University of North Carolina, Greensboro'
  - noaa: National Oceanic and Atmospheric Administration
abstract:
  Coral reef communities are influenced by various environmental and anthropogenic factors, yet those factors have not been consistently resolved. Furthermore, due to logistical difficulties, investigations of the role of depth (particularly mesophotic coral ecosystems) in influencing coral reef community structure have been taxonomically limited. Here, we examined the effects of a suite of environmental, ecological, and anthropogenic factors on the community structure of shallow brachyuran crabs across the ~2,500 km span of the Hawaiian Archipelago. In addition, we compared shallow crab communities with those sampled on a depth transect from 12--90 m on O&#x02BB;ahu. We deployed Autonomous Reef Monitoring Structures (ARMS) on shallow reefs (12--16 m) across the latitudinal gradient of the Hawaiian Archipelago to compare with a depth gradient (12, 30, 60, and 90 m) on the south shore of O&#x02BB;ahu. We found a total of 2,284 individual crabs representing 90 morphospecies and 19 families. Shallow crab communities were significantly different between the Main and Northwest Hawaiian Islands and variation in community structure was significantly associated with six biotic, environmental, and anthropogenic factors (cumulative human impacts, chlorophyll-A, sea-surface temperature, site depth, island slope, and potential larval connectivity). In depth comparisons, we found more variation among mesophotic samples from O&#x02BB;ahu (30--90 m) than among all shallow samples across the entire archipelago. We show that 90 m of depth is a more potent determinant of brachyuran community composition than the latitudinal, environmental, or anthropogenic gradients across the Hawaiian Islands.<!--Deeper reefs host significantly different brachyuran communities than shallow ones in Hawai&#x02BB;i, and only 3 of 90 morphospecies (~3%) occurring across the entire depth range (mesophotic to shallow) were sampled. Further research on the cryptic reef-dwelling fauna is needed for identifying areas of unique biodiversity, as well as boundaries between depth-stratified reef ecosystems. Shallow coral reefs are extensively studied, and although scleractinian corals have been recorded to ~170 m, little is known about other mesophotic coral reef ecosystem (MCE) inhabitants. Brachyuran crabs fill many ecological and trophic niches on reefs, making them ideal candidates for evaluating species composition among depths. Here we ask if MCEs host the same communities as the shallower reefs spread across the entire 2400 km Hawaiian Archipelago.-->
bibliography: references/citations.bib
csl: references/style.csl
output:
  word_document:
    pandoc_args:
      - "--toc-depth=2"
      - "--filter=resources/pandoc-crossref"
      - "--lua-filter=resources/scholarly-metadata.lua"
      - "--lua-filter=resources/author-info-blocks.lua"
      - "--reference-doc=resources/ref.docx"
    keep_md: true
tblPrefix: table
figPrefix:
  - "figure"
  - "figures"
chapters: True
chaptersDepth: 1
chapDelim: ""
header-includes:
  - \usepackage{bm}
  - \usepackage{upgreek}
---
<!--word count: Proceedings B publishes articles up to 12 pages, including figures and tables, with no more than 10 000 words. The word count includes the title page, author names, abstract, figure legends, tables and the reference list. As a rough guide, 10 000 words = 12 pages without additional figures; 850 words = 1 page and 2-4 small figures = 1 page. For example, an acceptable paper could be made up of 9 000 words and 4 small figures or 8500 words, 2 large figures and 4 small figures.-->
### Keywords {-}
<div custom-style="mono">ARMs, mesophotic coral ecosystems, community structure, dispersal, Hawai&#x02BB;i, brachyuran crabs</div>

@@

print_table <- function(tbl,id="",caption="",hdr_case="sentence",hide=character(0),accuracy = 0.01,format_numbers=TRUE) {
  if (!grepl("^: ",caption)) {
    caption <- paste0(": ",caption)
  }

@@

  tbl <- tbl %>%
    select(-any_of(hide)) %>%
    mutate(
      across(
        where(~format_numbers & is.numeric(.x)),
        ~num(.x,accuracy)
      )
    )

  # can also use rename_with(str_to_sentence)
  names(tbl) <- switch(
    str_to_lower(hdr_case),
    sentence = str_to_sentence(names(tbl)),
    title = str_to_title(names(tbl)),
    upper = str_to_upper(names(tbl)),
    lower = str_to_lower(names(tbl)),
    names(tbl)
  )

@@

island_pal <- c(
  "Hawaii Island" = "#5E4FA2",
  "Hawai‘i Island" = "#5E4FA2",
  "Maui" = "#3288BD",
  "Oahu" = "#66C2A5",
  "O‘ahu" = "#66C2A5",
  "Kauai" = "#ABDDA4",
  "Kaua‘i" = "#ABDDA4",

@@

The structure and composition of coral reef communities is influenced by a suite of environmental and anthropogenic factors, yet the nature and makeup of those factors remain poorly resolved [@hughesCommunityStructureDiversity1989; @smithEffectsTopBottom2010; @brandlCoralReefEcosystem2019]. Coral reefs are among the most diverse habitats in the world, and they face increasing threats of degradation, including climate-driven bleaching and phase shifts to alternate ecological regimes [@grahamManagingResilienceReverse2013; @norstromAlternativeStatesCoral2009; @pandolfiGlobalTrajectoriesLongTerm2003]. In the face of these threats, understanding drivers of metazoan community structure in coral reef ecosystems is critically important [@hughesRisingChallengeSustaining2010; @pageSeekingResistanceCoral2019], particularly as the search for criteria to prioritize climate refugia such as upwelling zone or mesophotic reefs intensifies. To date, the majority of such efforts have focused on conspicuous taxa such as corals and fishes [e.g., @benfieldComparisonCoralReef2008; @caballero-aragonWaveExposureTemperature2023; @kaneTrophicDesignationLive2017; @langeWaveExposureShapes2021; @vanlierHabitatConnectivityComplexity2018]. However, these taxa make up a minority of coral reef biodiversity and thus may not be indicative of community patterns overall [@fautinOverviewMarineBiodiversity2010]. Here, we investigate the factors influencing the community structure of brachyuran crabs across a suite of anthropogenic and environmental gradients using the Hawaiian Archipelago as a natural laboratory.

The Hawaiian Archipelago comprises high volcanic islands and atolls stretching across ~2,500 km in the North Pacific Ocean and is the most isolated island chain on the planet. Typically regarded as two regions, the main Hawaiian Islands (MHI) include the eight inhabited high islands, and the Northwestern Hawaiian Islands (NWHI) comprise the uninhabited islands, atolls, and banks northwest of Kaua&#x02BB;i and Ni&#x02BB;ihau that make up the Papahānaumokuākea Marine National Monument. The islands are subject to considerable anthropogenic and physical gradients across their span [@selkoeMapHumanImpacts2009], which are widely believed to influence coral reef communities. These gradients include differences in temperature, human population, and anthropogenic impact, among others. Studies examining community structure across broad environmental gradients have been criticized for confounding multiple potentially covarying factors (such as habitat area and/or complexity) across the measured gradients [e.g., @bellwoodRegionalScaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007a]. Thus, we used autonomous reef monitoring structures (ARMS)---standardized collection devices designed to mimic reef structural complexity---to sample cryptic communities while controlling for habitat [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009]. We then sampled brachyuran crabs, a group comprising a notable proportion of the coral reef cryptofauna [289 species in Hawai&#x02BB;i, @castroCatalogAnomuranBrachyuran2011] from these structures to examine drivers of community composition across these gradients. We compared crab communities among islands and regions (MHI vs. NWHI) and examined the effects of a suite of possible environmental and anthropogenic drivers of community structure to test hypotheses about their relative importance.

Much of our understanding of biotic zonation and community structure in the ocean originates with classic studies from the intertidal [e.g., @connellInfluenceInterspecificCompetition1961; @connellEffectsCompetitionPredation1961; @paineIntertidalCommunityStructure1974]. While there are no shortages of exceptions, a long history of studies point towards a generalization that the lower bathymetric limits are generally regulated by ecological interactions whereas upper limits of intertidal communities are more often determined by abiotic factors [@stephenson1972life; @mengeRockyIntertidalCommunities2001; @bertnessTestingRelativeContribution1999; @dennyPhysicalProcessesThat2001]. However, most of these models describe zonation at specific localities rather than community variation over larger spatial scales [but see @birdTideWaveMarks2013; @bishop-taylorTidesModellingElevation2019] and may not apply to coral reefs, which are subject to more complex community dynamics [@abramsRecruitmentLotteriesCoexistence1984; @bussCompetitiveNetworksNontransitive1979; @hixonPredationPreyRefuges1993; @hubbellUnifiedNeutralTheory2011; @moraEcologyFishesCoral2015; @saleStockRecruitRelationshipsRegional1982; @sheppardBiologyCoralReefs2017]. While there is little doubt that ecological interactions are important on coral reefs, the suite of factors and which play the most important role in structuring coral reef communities remains a subject of considerable debate [@bellwoodRegionalScaleAssemblyRules2001; @connollyUnifiedModelExplains2017; @hubbellUnifiedNeutralTheory2011; @jacksonAdaptationDiversityReef1991; @mumbyCoralReefHabitats2008; @saleMaintenanceHighDiversity1977].

One aspect of coral reef community structure that is thought to be important but remains under-investigatedis depth. Most studies to date have focused on the upper ~30 m of the water column, yet coral reefs and associated communities are known to extend beyond 150 m [@pyleExploringDeepCoral1996; @pyleMesophoticCoralEcosystems2019], with photosymbiotic corals reported to 170 m [@rouzeSymbioticAssociationsDeepest2021]. As much as two thirds of coral reef ecosystems lie in this under-explored zone [@hindersteinThemeSectionMesophotic2010]. These habitats, known as mesophotic coral ecosystems (MCEs), are typically defined as those extending from ~30--150 m depth [@hindersteinThemeSectionMesophotic2010; @kahngCommunityEcologyMesophotic2010]. They have been hypothesized as possible refugia from climate change impacts on shallow reefs [@glynnCoralReefBleaching1996; @bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although a growing number of studies have found evidence that mesophotic reefs are unique and uniquely threatened, challenging this notion [@diazMesophoticCoralBleaching2023; @diazMesophoticCoralBleaching2023; @hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018; @montgomeryCommunitySimilaritySpecies2021]. Given the strong reported differences between deep and shallow reefs across taxa and the paucity of research on drivers of cryptic coral reef community variation, we sought to answer the following four research questions: (i) How do shallow crab communities vary along the natural and anthropogenic gradients of the Hawaiian Archipelago? (ii) Do shallow crab communities differ between the Main and Northwestern Hawaiian Islands? (iii) What environmental and anthropogenic factors drive crab community structure in shallow coral reefs across the Hawaiian Archipelago? (iv) What are the relative contributions of depth (shallow vs. mesophotic) and island region to crab community variation across the Hawaiian Archipelago?

@@

ARMS units were retrieved by encapsulating them in-situ in a 100 μm mesh-lined crate to prevent escape of motile organisms. Once returned to the surface, they were placed in containers of aerated, filtered seawater, crates were removed, and units were systematically disassembled plate-by-plate [see @lerayDNABarcodingMetabarcoding2015 for processing details]. All adult brachyuran crabs ($\geq$ 5 mm) were photographed, identified (to species where possible, otherwise morphospecies), and preserved in 95% ethanol.<!-- Specimens were collected under permits #xxx (NWHI) and #xxx (MHI) and accessioned into the invertebrate zoology collections at the Florida Museum of Natural History (catalog numbers xxx--xxx).-->

@@

To examine patterns of crab diversity across the islands, we used `estimate_richness` to calculate species richness and Simpson diversity (Simpson = $1-\sum_{i=1}^{R}p_i^2$, where R is the total number of species, and $\mathrm{p}_i$ is the proportional abundance of each species). We used nonparametric Kruskal-Wallis ranked sum tests to compare diversity between shallow and mesophotic samples and between island groups (MHI vs. NWHI). Island group comparisons were performed both including and excluding mesophotic samples. <!--To explore environmental influences on alpha diversity, we used a linear mixed-effect model with Simpson or richness as response variables, the environmental variables listed below as fixed effects, and island nested within island group (MHI vs NWHI) as random effects. Models were generated using the `lmer` function from the `lme4` package [@batesFittingLinearMixedEffects2015]. -->

@@

We assessed variation among groups using `betadisper` and `adonis2`. These functions implement analysis of multivariate homogeneity of group dispersions (PERMDISP) and multivariate permutational analysis of variance (PERMANOVA) respectively. PERMDISP is used to assess variability within groups and PERMANOVA is used to assess differences among groups. Significance was assessed with permutation tests in both cases. Variation among sampling groups (island group and sampling depth) was further examined and visualized using canonical analysis of principal coordinates (CAP) based on discriminant analysis [@andersonCanonicalAnalysisPrincipal2003] as implemented in the `CAPdiscrim` function in `BiodiversityR` [@kindtTreeDiversityAnalysis2005].

@@

The `r nsamples(cc$crabs_untransformed)` ARMS units deployed across the Hawaiian Islands (in `r format_range(cc$crab_data$depth)` m depth) yielded a total of `r num(sum(sample_sums(cc$crabs_untransformed)),1)` adult crabs. Units contained `r num(all_summary$abundance_mean,0.1)` $\pm$ `r num(all_summary$abundance_sd,0.1)` (mean $\pm$ SD) individuals, and abundance ranged from `r format_range(all_summary$abundance_range)`. Crabs were identified to `r sum(unlist(all_summary$types_breakdown))` distinct types (`r all_summary$types_breakdown$species` species and `r all_summary$types_breakdown$morphospecies` morphospecies) in `r all_summary$taxonomy_breakdown$genera` genera and `r all_summary$taxonomy_breakdown$families` families, which accounted for ~36% of the known crab diversity reported from Hawaiian waters [@castroCatalogAnomuranBrachyuran2011]. The four most abundant species were: `r with(all_summary,top_list(top,top_n,top_units,fig="fig_crabs"))`. There were `r all_summary$singleton_count` singleton taxa (morpho/species occurring in a single unit across all sites), accounting for ~`r num(all_summary$singleton_count/ntaxa(cc$crabs_untransformed)*100,1)`% of total recovered crab species. <!--`r str_to_sentence(english(all_summary$singleton_count))` species occurred in a single unit across all sites (singletons), accounting for ~`r num(all_summary$singleton_count/ntaxa(cc$crabs_untransformed)*100,1)`% of total recovered crab species.-->

@@

The `r nsamples(cc$crabs_deep)` units deployed at mesophotic depths $\geq$ 30 m on O&#x02BB;ahu yielded fewer species and individuals compared to the complete dataset, with `r num(sum(sample_sums(cc$crabs_deep)),1)` crabs overall, `r sum(unlist(deep_summary$types_breakdown))` distinct types (`r deep_summary$types_breakdown$species` species and `r deep_summary$types_breakdown$morphospecies` morphospecies), `r num(deep_summary$abundance_mean,0.1)` $\pm$ `r num(deep_summary$abundance_sd,0.1)` individuals per unit, and abundance ranging from `r format_range(deep_summary$abundance_range)`. The four most abundant taxa in the deep ARMS did not overlap with those from the complete dataset: `r with(deep_summary,top_list(top,top_n,top_units))`. There were `r deep_summary$singleton_count` singletons, which proportionally accounted for ~`r num(deep_summary$singleton_count/ntaxa(cc$crabs_deep)*100,1)`% of species found on deep ARMS.

@@

<!-- ### Effects of environmental variables {-} -->
## Brachyuran community structure {-}
